---
knit: "bookdown::preview_chapter"
editor_options: 
  chunk_output_type: console
---

# Condensing data with numerical summaries {#chapter2}

> James Bond: [to Vesper] Why is it that people who can't take advice always insist on giving it? 

<!-- (90 minutes) -->

One of the roles of a data scientist is to condense large amounts of information
into a few key summaries. There are many possible choices when simplifying data
and it is certainly not the case that "one size fits all".
In this chapter, we will look at standard measures of location and spread, which are
the building blocks for many statistical and machine learning algorithms. We discover
that some measures fit naturally into a big data scenario while others do not.


```{r echo=FALSE}
source("code/load_data.R")
```

## Mathematical notation

Before we can talk more about numerical techniques we first need to
define some basic notation. This will allow us to generalise all
situations with a simple shorthand.

Very often in statistics we replace actual numbers with letters in
order to be able to write general formulae. As discussed earlier, we generally use a single upper case letter to represent our random variable and the lower case to represent sample data, with subscripts to distinguish individual observations in the sample.  Amongst the most common letters to use is $x$, although $y$ and $z$ are frequently used as well. For
example, suppose we ask a random sample of three people how many
mobile phone calls they made yesterday. We might get the following data: 1, 5, 7. If
we take another sample we will most likely get different data, say 2,
0, 3. Using algebra we can represent the general case as $x_1$, $x_2$,
$x_3$:

| | |
------------|--|---|---
1st sample: | 1 | 5 | 7
2nd sample: | 2 | 0 | 3 
$\vdots$  | $\vdots$ | $\vdots$ | $\vdots$ 
typical sample: | $x_1$ | $x_2$ | $x_3$

This can be generalised further by referring to the random variable _as a whole_ as
$X$ and the $i\text{th}$ observation in the sample as $x_i$. Hence, in
the first sample above, the second observation is $x_2=5$
whilst in the second sample it is $x_2=0$. The letters $i$ and $j$ are
most commonly used as the index numbers for the subscripts.

The total number of observations in a sample, that is the _sample size_, is usually referred to by the letter $n$.
Hence in our simple example above $n = 3$.

The next important piece of notation to introduce is the symbol
$\sum$. This is the upper case of the Greek letter $\sigma$, pronounced ''sigma''.  It is used
to represent the phrase 'sum the values'. This symbol is used as
follows
\[
\sum_{i=1}^n x_i = x_1 + x_2 + \dots + x_n.
\]
This notation is used to represent the sum of all the values in our
data (from the first $i = 1$ to the last $i = n$), and is often
abbreviated to $\sum x$ when we sum over all the data in our sample.

## Measures of location

These are also referred to as averages. In general
terms, they tell us the value of a "typical" observation. There are two measures which are
commonly used: the mean and the median. We will consider these in
turn. When these are obtained from a sample of data, rather than the entire population, we
often prefix with the word "sample".

### Sample mean

One of the most important and widely used measures of location is the (arithmetic) mean,
more commonly known as the average
\[
 \bar x = \frac{x_1 + x_2 + \ldots + x_n}{n} = \frac{1}{n} \sum_{i=1}^n x_i 
\]
where $x_i$ are our data points and $n$ is the sample size, i.e. the number of 
data points in our sample. So if our data set was $\{0,3,2,0\}$, then $n=4$. Hence,
\[
\bar x = \frac{1}{n} \sum_{i=1}^n x_i = \frac{0 + 3 + 2 + 0}{4} = 1.25 \;.
\] 
In statistics, it is common to use a potential bar over a variable to denote the 
mean.


```{block type="rmdnote"}
Why do we say "sample mean" instead of just mean? In statistics we make the
distinction between the population mean (the _true_ value) and the estimate mean.
To differentiate between the population and sample means, we use $\mu$ and $\bar x$
respectively. 

Statistics uses the sample to inform us about the population.
```

### Example: The beauty data set {-}

The `beauty` data set contains results for `r nrow(beauty)` classes. After loading 
the data set into R, we can extract a particular column using the dollar 
operator
```{r, results="hide"}
## Attractiveness score
beauty$beauty
## Number of students per class
beauty$students
```
We can use the built-in function `mean()` to calculate the
average attractiveness score and class size
```{r}
## Attractiveness score (normalised)
mean(beauty$beauty)
## Number of students per class
mean(beauty$students)
```

### Sample median

The median is occasionally used instead of the mean, particularly when the data have an
asymmetric profile or when there are outliers (as indicated by graph, perhaps). The median is
the middle value of the observations when they are listed in ascending order. It is
straightforward to determine the median for small data sets. For larger data sets, the calculation
is more easily done using a computer package such as R.


Ranked observations are denoted as $x_{(1)}, x_{(2)}, \ldots, x_{(n)}$. The sample median is defined as:
\[
\text{Sample median} = 
\begin{cases}
  x_{(n+1)/2}, & n \text{ odd;} \\
  \frac{1}{2}x_{(n/2)} + \frac{1}{2} x_{(n/2+1)}, & n \text{ even}\;.
\end{cases}
\]

The median is less sensitive to outliers than the sample mean, but has less useful mathematical properties. 

For our simple data set $\{0,3,2,0\}$, to calculate the median we
re-order it to: $\{0,0,2,3\}$; then take the average of the middle two
observations, to get 1.

#### Example: The beauty data set {-}

R has a built-in function called `median()` that we can use to calculate sample median.
We'll use the same variables as previous
```{r}
median(beauty$beauty)
median(beauty$students)
```
Notice that the mean and median beauty scores are approximately the same; this implies that the data
are roughly symmetric. However the class size values differ: $\bar x = 55$ and the median is $29$. This
is because the class sizes are [skewed](https://en.wikipedia.org/wiki/Skewness); 
there are many more large values than smaller values.

```{block type="rmdnote"}
The classic case of where it can be misleading to use the mean, is income. Here
a few very rich individuals can dominate the statistic.
```

## Measures of spread
<!-- (25 minutes) -->


A measure of location is insufficient in itself to summarise data as it only describes the value of a
typical outcome and not how much variation there is in the data. For example, the two datasets
10, 20, 30 and 19, 20, 21 both have the same mean (20) and the same median (20). However the
first set of data ranges considerably from this value while the second stays very close. They are
quite clearly very different data sets. The mean or the median does not fully represent the data.
There are three basic measures of spread which we will consider: the range, the inter–quartile
range and the sample variance/standard deviation.
  
### Range
  
The range is easy to calculate. It is simply the largest minus the
smallest.^[When you get a new data set, calculating the range is
    useful when checking for obvious data-input errors.]
\[
\text{Range }= x_{(n)} - x_{(1)} 
\]
where $x_{(n)}$ is the largest value in our data set and $x_{(1)}$ is the smallest value.
So for our data set of $\{0,3,2,0\}$, the range is $3-0=3$. It is very useful
for data checking purposes, but in general it's not very robust. 

There are two problems with the range as a measure of spread. When calculating the range you
are looking at the two most extreme points in the data, and hence the value of the range can be
unduly influenced by one particularly large or small value, known as an outlier. The second
problem is that the range is only really suitable for comparing (roughly) equally sized samples as
it is more likely that large samples contain the extreme values of a population.


#### Example: OKCupid {-}

Starting any analysis it's a good idea to calculate the range of the variables. 
It's a straightforward way of ensuring everything is correct.
For example, the range of ages in the OK Cupid data set
is
```{r}
range(cupid$age)
```
Perhaps a little on the high side?


### Sample variance and standard deviation
  
In statistics, the mean and variance are used most often. 
This is primarily because they have nice mathematical properties, unlike the median. 
The sample variance, $s^2$, is defined as
\[
s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2
\]
Essentially we are measuring the difference between the sample mean, $\bar x$,
and each observation in turn. By squaring the difference, we avoid positive and negative
values cancelling.

The formula can be rewritten as
\[
s^2= \frac{1}{(n-1)} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2 \right\} \;.
\]
The second formula is easier for calculations.^[The divisor is $n-1$ rather than
  $n$ in order to correct for the bias which occurs because we are measuring
deviations from the sample mean rather than the ``true" mean of the population
  we are sampling from.] However $\sum x^2$ becomes very large as $n$ increases.
This could lead to a potential loss of precision, in practice we would 
use
\[
s^2= \sum_{i=1}^n \frac{x_i^2}{n-1}  - \left(\frac{n}{n-1}\right) \bar x^2  \;.
\]  
For our toy data set, $\{0,3,2,0\}$, we have
\[
\sum_{i=1}^4 x_i^2 = 0^2 + 3^2 + 2^2 + 0^2 = 13.
\]
So
\[
s^2 = \frac{1}{n-1} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2
\right\} = \frac{1}{3} \left( 13 - 4 \times 1.25^2\right) = 2.25.
\]
The sample standard deviation, $s$, is the square root of the sample variance,
i.e. for our toy example $s = \sqrt{2.25} = 1.5$. The standard
deviation is preferred as a summary measure as it is in the same units as the
original data. However, it is often easier from a theoretical perspective to
work with variances.

A rule of thumb for getting a feel for your data, is that approximately 95% of data is contained
in the interval
\[
\bar x \pm 2 \times s
\]
We'll see in chapter \@ref(chapter4) the mathematical underpinning for this rule.

```{block, type="rmdnote"}
If this was a "Maths class", we would spend some time discussing why we
divide by $n - 1$ and not $n$. If we had access to all of the population, we would use $n$. 
However, we don't and to correct for that we use $n - 1$. 
```

#### Example: R

R has built-in functions to calculate the standard deviation and variance respectively
```{r}
sd(beauty$evaluation)
var(beauty$evaluation)
```

### Quartiles and the interquartile range
  
The upper and lower quartiles are defined as follows:

  * $Q_1$: Lower quartile = ($n$+1)/4$^{\text{th}}$ smallest observation.
  * $Q_3$: Upper quartile = 3 ($n$+1)/4$^{\text{th}}$ smallest observation.

We can linearly interpolate between adjacent observations if necessary. The 
interquartile range is the difference between the third and first quartile, i.e.
\[
IQR = Q_3 - Q_1.
\]
Like the median, the IQR isn't influenced by large values (outliers). However it
is more computationally difficult to calculate.

### Numerical examples
  
For the following data sets, calculate the inter-quartile range:

1. $\{5, 6, 7, 8\}$ 

The lower quartile is the $5/4 = 1.25^{\text{th}}$ smallest
  observation, i.e. $x_{(1.25)}$. The value of $x_{(1)} = 5$ and $x_{(2)} = 6$. So the first quartile is one quarter of the way between $x_{(1)}$ and $x_{(2)}$, i.e a quarter of the way between 5 and 6,
  \[
  x_{(1.25)} = x_{(1)} + 0.25 \times (x_{(2)} - x_{(1)}) = 5 + \frac{6-5}{4} = 5.25.
  \]
  Similarly, the upper quartile is the $3 \times (4+1)/4 = 3.75^{\text{th}}$ smallest observation. So 
  \[
  x_{(3.75)}  = x_{(3)} + 0.75 \times (x_{(4)} - x_{(3)}) = 7.75.
  \]
  Therefore, 
  \[
  IQR = 7.75 - 5.25 = 2.5.
  \]
  
2. $\{10, 15, 20, 25, 50\}$. 

The lower quartile is the $(5+1)/4 = 1.5^{\text{th}}$ smallest
  observation, i.e. $x_{1.5}$, so
  \[
  x_{(1.5)} = x_{(1)} + 0.5 \times (x_{(2)} - x_{(1)}) = 12.5.
  \]
  Similarly, the upper quartile is the $3 \times (5+1)/4 = 4.5^{\text{th}}$ smallest observation. So 
  \[
  x_{(4.5)} = x_{(4)} + 0.5 \times (x_{(5)}-x_{(4)}) = 37.5.
  \]
  Thus,
  \[
  IQR = 37.5 - 12.5 = 15.
  \]
  
#### R example

The quartiles are special cases of percentiles: 

 * The minimum is the 0$^{th}$ percentile
 * $Q_1$ is the 25$^{th}$ percentile
 * The median, $Q_2$ is the 50$^{th}$ percentile
 * $Q_3$ is the 75$^{th}$ percentile
 * The maximum is the 100$^{th}$ percentile

The `quantile()` function calculate these values
```{r}
quantile(cupid$age)
```
The R function has also has a `probs` argument where we can specify our
own percentiles
```{r}
## The middle 90% 
quantile(cupid$age, probs = c(0.05, 0.95))
```

```{block, type="rmdtip"}
There are actually (at least) 9 different versions of how to calculate
quantiles (see `?quantile` in R). The version described in these notes is `type = 6` in R.
Typically, the quantiles give approximately the same answer. 
```

<!-- (25 minutes) -->

## Streaming data

Streaming data is data that is generated continuously by multiple sources. 
Developing algorithms for dealing with streaming data is an active area of research,
both statistically and computationally. Even though we've only covered a few statistics,
some of them can be easily adapted to deal with streaming data while others can't.

### The mean and variance

Suppose we have observed $k-1$ values and our current estimate of the mean is $m_{k-1}$.
A new observation, $x_k$ arrives. How would we update our estimate? 
An obvious method is
\[
m_k = \frac{(k-1) \times m_{k - 1} + x_k}{k}
\]
However this method isn't _numerically stable_. Basically $(k-1) \times m_{k - 1}$ can 
get large and we'll lose precision. Instead we should use
\[
m_k = m_{k-1} + \frac{x_k - m_{k-1}}{k}
\]
We have a similar algorithm for the variance
\[
v_k = v_{k-1} + (x_k - m_{k-1})(x_k - m_k)
\]
Again this algorithm is [numerically stable](https://en.wikipedia.org/wiki/Numerical_stability).

#### Example: R 

The OKCupid data set contains almost 60,000 individuals. Figure \@ref(fig:2-1)
plots the estimate of the mean age as we progress through the data set.
```{r, 2-1, echo=FALSE, fig.cap = "The estimate of the mean age as we progress through the data set."}
local(source("code/f2_average_age.R"))
```
After some variation in the first few observations, we quickly settle to a consistent
estimate of the mean. The spike up to age $34$ is due to the outlying age of $110$

### The median and quantiles

In contrast to the mean keeping a [running median](https://stackoverflow.com/q/10657503/203420)
is a non-trivial task. Essentially we need to maintain a sorted data structure containing 
the data. The key issues are storage cost of the data and retrieval time. 

## Relevant R functions {-}

```{r echo=FALSE}
library(tibble)
r_functions = tribble(
  ~Command, ~Comment, ~Example, 
"mean", "Calculates the mean of a vector", "`mean(x)`",
"sd", "Calculates the standard deviation of a vector", "`sd(x)`",
"var", "Calculates the variance of a vector", "`var(x)`",
"quantile", "The vector quartiles." , "`quantile(x, type = 6)`",
"range", "Calculates the vector range", "`range(x)`"
)  

knitr::kable(
  r_functions, longtable = TRUE, booktabs = TRUE,
  caption = 'Summary of R commands in this chapter.'
)
```

    
