---
title: "Condensing data with numerical summaries"
subtitle: "Chapter 2"
author: '`r jrPresentation::get_author()`'
output:
  xaringan::moon_reader:
    css: ["default", "style.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library("jrPresentation")
set_presentation_options()
```


layout: true
`r add_border(inverse=TRUE)`
---
class: inverse, center, middle

> James Bond: [to Vesper] Why is it that people who can't take advice always insist on giving it? 

---
layout: true
`r add_border(inverse=FALSE)`

---
```{r echo=FALSE, cache=FALSE}
source("code/load_data.R")
```

# Overview

  * Data scientists need to condense large amounts of information into a few key summaries
  
  * There are many possible choices when simplifying data
  
    - not "one size fits all".

  * Look at standard measures of location and spread
  
    - building blocks for many statistical and machine learning algorithms
    
  *  We discover that some measures fit naturally into a big data scenario while others do not

---

# A wee bit of maths

  * Basic notation
    
    - This will allow us to generalise all situations with a simple shorthand.
    
  * Very often in statistics we replace actual numbers with letters in order to be able to write general formulae. 
  
    - Upper case letter to represent our random variable 
    - Lower case to represent sample data, with subscripts to distinguish individual observations in the sample.
    
---

# Example


  * Suppose we ask three people how many mobile phone calls they made yesterday.
  
    - We might get the following data: 1, 5, 7.
    - Another sample we will most likely get different data, say 2, 0, 3. 
    - Algebra represents the general case as $x_1$, $x_2$, $x_3$:

--
| | |
------------|--|---|---
1st sample: | 1 | 5 | 7
2nd sample: | 2 | 0 | 3 
$\vdots$  | $\vdots$ | $\vdots$ | $\vdots$ 
typical sample: | $x_1$ | $x_2$ | $x_3$

---

# Further generalisation

  * Random variable _as a whole_ as $X$ and the $i\text{th}$ observation in the sample as $x_i$.
  
  * Previously, we had 1, 5, 7 so
  * $x_1 = 1$, $x_2 = 5$ and $x_3 = 7$
The total number of observations in a sample, that is the _sample size_, is usually referred to by the letter $n$.
Hence in our simple example above $n = 3$.

---

# Sums

  * The next important piece of notation to introduce is the symbol $\sum$
    - Upper case of the Greek letter $\sigma$, pronounced ''sigma''.  
    
--

  * It is used to represent the phrase 'sum the values'. 

$$
\sum_{i=1}^n x_i = x_1 + x_2 + \dots + x_n.
$$

  - Sum of all the values in our data (from the first $i = 1$ to the last $i = n$)
  
  - Often shorten to $\sum x$

---

# Measures of location (averages)

  - Whats a typical observation
    - mean and the median.

---

# Sample mean


$$\bar x = \frac{x_1 + x_2 + \ldots + x_n}{n} = \frac{1}{n} \sum_{i=1}^n x_i$$


where $x_i$ are our data points and $n$ is the sample size, i.e. the number of 
data points in our sample

---

# Example

So if our data set was $\{0,3,2,0\}$, then $n=4$. Hence,
$$\bar x = \frac{1}{n} \sum_{i=1}^n x_i = \frac{0 + 3 + 2 + 0}{4} = 1.25 $$

In statistics, it is common to use a potential bar over a variable to denote the 
mean.

---

# Example: The beauty data set

  * This is __not__ made up
  
  * The `beauty` data set contains results for `r nrow(beauty)` classes
  
  * Does attractiveness affect teaching ability
  
---

# Example: The beauty data set

After loading the data set into R we can extract a particular column using the dollar 
operator
```{r, results="hide"}
## Attractiveness score
beauty$beauty
## Number of students per class
beauty$students
```

---

# Example: The beauty data set

We can use the built-in function `mean()` 
```{r}
## Attractiveness score (normalised)
mean(beauty$beauty)
## Number of students per class
mean(beauty$students)
```

---

# Sample median

  * Occasionally used instead of the mean, 
    - Particularly when the data have an asymmetric profile or when there are outliers (as indicated by graph, perhaps). 
    - Wages!
  * Work with _ranked_ observations
    - $x_{(1)}, x_{(2)}, \ldots, x_{(n)}$
    - $x_{(1)}$ minimum
    - $x_{(n)}$ maximum

---

# Sample median

* Odd sample size
$$\text{Sample median} =x_{(n+1)/2}$$
 * Even sample size

$$\text{Sample median} =\frac{1}{2}x_{(n/2)} + \frac{1}{2} x_{(n/2+1)}$$

---

# Example example


For our simple data set $\{0, 3, 2, 0\}$, to calculate the median we

  - re-order it to: $\{0, 0, 2, 3\}$
  - take the average of the middle two observations, to get 1.

---

# Example: The beauty data set 

R has a built-in function called `median()` that we can use to calculate sample median.
We'll use the same variables as previous
```{r}
median(beauty$beauty)
median(beauty$students)
```


---
layout: true
`r add_border(inverse=TRUE)`
---
class: inverse, middle, center

# Measures of spread

---
layout: true
`r add_border(inverse=FALSE)`
---

# Measures of spread

A measure of location is insufficient in itself to summarise data as it only describes the value of a
typical outcome and not how much variation there is in the data. 
  
  * Compare:
    * 10, 20, 30 and 19, 20, 21 
    * both have the same mean (20) and the same median (20)
    * both are clearly different

---

# Range
  
The range is easy to calculate: largest minus the smallest

$$\text{Range }= x_{(n)} - x_{(1)}$$

where $x_{(n)}$ is the largest value in our data set and $x_{(1)}$ is the smallest value.

--

So for our data set of $\{0,3,2,0\}$, the range is $3-0=3$. It is very useful
for data checking purposes, but in general it's not very robust. 

---

# Range: issues

1. Unduly influenced by one particularly large or small value, known as an outlier. 
1. Only really suitable for comparing (roughly) equally sized samples as
it is more likely that large samples contain the extreme values of a population
--

> Useful for data checking

---

# Example: OKCupid ranges


```{r}
range(cupid$age)
```

Perhaps a little on the high side?

---

# Sample variance and standard deviation
  
  * The mean and variance are the most popular
  * Nice mathematical properties, unlike the median. 
  
The sample variance, $s^2$:

$$s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2$$

--

The formula can be rewritten as

$$s^2= \frac{1}{(n-1)} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2 \right\}$$


---

# Sample variance and standard deviation
  

  * However $\sum x^2$ becomes very large as $n$ increases.
  
    - Loss of precision
    
  * In practice we would use
$$s^2= \sum_{i=1}^n \frac{x_i^2}{n-1}  - \left(\frac{n}{n-1}\right) \bar x^2$$


---

# Example

For our toy data set, $\{0,3,2,0\}$, we have

$$\sum_{i=1}^4 x_i^2 = 0^2 + 3^2 + 2^2 + 0^2 = 13$$

So
$$s^2 = \frac{1}{n-1} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2 \right\} = \frac{1}{3} \left( 13 - 4 \times 1.25^2\right) = 2.25$$

 * The sample standard deviation, $s$, is the square root of the sample variance, 
  - $s = \sqrt{2.25} = 1.5$. 

  * Standard deviation is in the same units as the
original data. 

---

# Rule of thumb

Approximately 95% of data is contained in the interval

$$\bar x \pm 2 \times s$$

---

# Example: Beauty


```{r}
sd(beauty$evaluation)
var(beauty$evaluation)
```

---

# Quartiles and the interquartile range
  
  * $Q_1$: Lower quartile = ($n$+1)/4$^{\text{th}}$ smallest observation.
  * $Q_3$: Upper quartile = 3 ($n$+1)/4$^{\text{th}}$ smallest observation.

$$IQR = Q_3 - Q_1$$

---

# R example

The quartiles are special cases of percentiles: 

 * The minimum is the 0$^{th}$ percentile
 * $Q_1$ is the 25$^{th}$ percentile
 * The median, $Q_2$ is the 50$^{th}$ percentile
 * $Q_3$ is the 75$^{th}$ percentile
 * The maximum is the 100$^{th}$ percentile

```{r}
quantile(cupid$age)
```

---

# R example

```{r}
## The middle 90% 
quantile(cupid$age, probs = c(0.025, 0.975))
```

--

```{r}
# Rule of thumb
mean(cupid$age) - 2*sd(cupid$age)
mean(cupid$age) + 2*sd(cupid$age)
```

---
class: inverse

# Streaming data

> Streaming data is data that is generated continuously by multiple sources. 

---

# The mean and variance

 * Suppose we have observed $k-1$ values and our current estimate of the mean is $m_{k-1}$.

--

  * A new observation, $x_k$ arrives. How would we update our estimate? 
--

An obvious method is
$$m_k = \frac{(k-1) \times m_{k - 1} + x_k}{k}$$

--

 * However this method isn't _numerically stable_. 

Basically $(k-1) \times m_{k - 1}$ can 
get large and we'll lose precision. 

---

# Streaming data: mean & variance

Instead we should use

$$m_k = m_{k-1} + \frac{x_k - m_{k-1}}{k}$$

We have a similar algorithm for the variance

$$v_k = v_{k-1} + (x_k - m_{k-1})(x_k - m_k)$$

Again this algorithm is [numerically stable](https://en.wikipedia.org/wiki/Numerical_stability).

---

# Example: OKCupid

The OKCupid data set contains almost 60,000 individuals. 

```{r, echo=FALSE}
local(source("code/f2_average_age.R"))
```

---

# The median and quantiles

  * Keeping a [running median](https://stackoverflow.com/q/10657503/203420) is a non-trivial task
  
  * Essentially we need to maintain a sorted data structure containing the data
  
  * The key issues are storage cost of the data and retrieval time. 

---

# Relevant R functions 

```{r echo=FALSE, results='asis'}
library(tibble)
r_functions = tribble(
  ~Command, ~Comment, ~Example, 
"mean", "Calculates the mean of a vector", "mean(x)",
"sd", "Calculates the standard deviation of a vector", "sd(x)",
"var", "Calculates the variance of a vector", "var(x)",
"quantile", "The vector quartiles." , "quantile(x, type = 6)",
"range", "Calculates the vector range", "range(x)"
)  

knitr::kable(
  r_functions, longtable = TRUE, booktabs = TRUE,
  caption = 'Summary of R commands in this chapter.', format = "html"
)
```

    
