---
title: "Condensing data with numerical summaries"
subtitle: "Chapter 2"
author: '`r jrPresentation::get_author()`'
output:
  xaringan::moon_reader:
    css: ["default", "style.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message=FALSE}
options(htmltools.dir.version = FALSE)
library("jrPresentation")
set_presentation_options()
```


`r add_border(inverse=TRUE, background_image = "assets/white_logo.png")`
---
class: inverse, center, middle

> James Bond: [to Vesper] Why is it that people who can't take advice always insist on giving it? 

---
`r add_border(inverse=FALSE)`

---
```{r echo=FALSE, cache=FALSE}
source("code/load_data.R")
```

# Overview

  * Data scientists need to condense large amounts of information into a few key summaries
  
  * There are many possible choices when simplifying data
  
    - "One size __doesn't__ fits all"

--

  * Look at standard measures of location and spread
  
    - Building blocks for many statistical and machine learning algorithms

--

  *  Some measures fit naturally into a big data scenario while others do not

---

# A [wee](http://www.dsl.ac.uk/entry/snd/wee_n1_adj_adv) bit of maths

  * Basic notation
    
    - Generalise all situations with a simple shorthand
    
  * We replace actual numbers with letters in order to be able to write general formulae
  
    - Upper case letter to represent our random variable 
    
    - Lower case to represent sample data
      - Subscripts to distinguish individual observations in the sample
    
---

# Example


Suppose we ask three people how many mobile phone calls they made yesterday
  
  - We might get the following data: 1, 5, 7
  
  - Another sample we will most likely get different data, say 2, 0, 3
    
  - Algebra represents the general case as $x_1$, $x_2$, $x_3$

--

| | | |
------------|--|---|---|---
1st sample: | 1 | 5 | 7
2nd sample: | 2 | 0 | 3 
$\vdots$  | $\vdots$ | $\vdots$ | $\vdots$ 
Typical sample: | $x_1$ | $x_2$ | $x_3$

---

# Further generalisation

  * A random variable as $X$ and the $i\text{th}$ observation in the sample as $x_i$.
  
  * Previously, we had 1, 5, 7 so
  
    * $x_1 = 1$, $x_2 = 5$ and $x_3 = 7$
--

  * The total number of observations in a sample $i$ the _sample size_
  
    - Referred to by the letter $n$; so $n = 3$ above

---

# Sums

  * The next important piece of notation to introduce is the symbol $\sum$
  
    - Upper case of the Greek letter $\sigma$, pronounced ''sigma''


  * It is used to represent the phrase 'sum the values'

$$
\sum_{i=1}^n x_i = x_1 + x_2 + \dots + x_n.
$$

  - Sum of all the values in our data (from the first $i = 1$ to the last $i = n$)
  
--

  - Often shorten to $\sum x$

---

# Measures of location (averages)

  - What's a typical observation
  
--

  - Mean and the median

---

# Sample mean


$$\bar x = \frac{x_1 + x_2 + \ldots + x_n}{n} = \frac{1}{n} \sum_{i=1}^n x_i$$


where 

  * $x_i$ are our data points
  * $n$ is the sample size, i.e. the number of data points in our sample

---

# Example

So if our data set was $0,3,2,0$, then $n=4$. Hence,
$$\bar x = \frac{1}{n} \sum_{i=1}^n x_i = \frac{0 + 3 + 2 + 0}{4} = 1.25 $$

In statistics, it is common to use a potential bar over a variable to denote the 
mean

---

# Example: The beauty data set

  * This is __not__ made up

---

# The relationship between beauty and teaching

[Study](https://doi.org/10.3386/w9853) where researchers were 
interested in whether a lecturers' attractiveness affected 
their course evaluation! 

--

 * evaluation: the questionnaire result
 
 * tenured: does the lecturer have tenure; 1 == Yes. In R, this value is continuous
 
 * minority: does the lecturer come from an ethnic minority (in the USA)

--

 * age: the lecturers' age
 
 * gender: a factor: Female or Male
 
 * students: number of students in the class
 
 * beauty: each of the lecturers' pictures was rated by six undergraduate students: three women and three men
 
---

# The relationship between beauty and teaching

  * The `beauty` data set contains results for `r nrow(beauty)` classes
  
  * Does attractiveness affect teaching ability?
  
---

# Example: The beauty data set

After loading the data set into R we can extract a particular column using the dollar 
operator
```{r, results="hide"}
## Attractiveness score
beauty$beauty
## Number of students per class
beauty$students
```

---

# Example: The beauty data set

We can use the built-in function `mean()` 
```{r}
## Attractiveness score (normalised)
mean(beauty$beauty)
```
--
```{r}
## Number of students per class
mean(beauty$students)
```

---

# Sample median

  * Occasionally used instead of the mean
  
    - Particularly when the data have an asymmetric profile or when there are outliers (as indicated by graph, perhaps)
    
    - Wages!
--

  * Work with _ranked_ observations
--

    - $x_{(1)}, x_{(2)}, \ldots, x_{(n)}$  - notice the brackets __()__
    - $x_{(1)}$ minimum
    - $x_{(n)}$ maximum

---

# Sample median

  * Odd sample size


$$\text{Sample median} =x_{(n+1)/2}$$
--

  * Even sample size

$$\text{Sample median} =\frac{1}{2}x_{(n/2)} + \frac{1}{2} x_{(n/2+1)}$$

---

# Example example


For our simple data set $\{0, 3, 2, 0\}$, to calculate the median we

  - re-order it to: $\{0, 0, 2, 3\}$

--
  
  - take the average of the middle two observations $(0 + 2)/2$, to get 1

---

# Example: The beauty data set 

R has a built-in function called `median()`

```{r}
median(beauty$beauty)
median(beauty$students)
```

---

# Exercise (10 minute or so)

For the beauty data set, calculate the `mean()` and `median()` for the
columns

  * tenured
  
  * minority
  
  * age

---
`r add_border(inverse=TRUE, background_image = "assets/white_logo.png")`
---
`r add_class(inverse=TRUE, middle=TRUE, center=TRUE)`

# Measures of spread

---
`r add_border(inverse=FALSE)`
---

# Measures of spread

A measure of location is insufficient in itself to summarise data 

  * It only describes the value of a typical outcome

--

  * Compare:

    * 10, 20, 30 and 19, 20, 21 

    * Both have the same mean (20) and the same median (20)

    * Both are clearly different

---

# Range
  
The range is easy to calculate - largest minus the smallest

--

$$\text{Range }= x_{(n)} - x_{(1)}$$

where $x_{(n)}$ is the largest value in our data set and $x_{(1)}$ is the smallest value.

--

So for our data set of $\{0,3,2,0\}$:

  * the range is $3-0=3$
  
  * Useful for data checking purposes

---

# Range: issues

1. Unduly influenced by one particularly large or small value, known as an outlier

--

1. Only really suitable for comparing (roughly) equally sized samples as 
it is more likely that large samples contain the extreme values of a population

--

> Useful for data checking

---

# Example: OKCupid ranges


```{r}
range(cupid$age)
```

Perhaps a little on the high side?

--

It turns out that there are:

  * 31 sixty nine year olds in the sample
  * A single person at 109
  * A single person at 110

---

# Exercise: 5 minute thought exercise

In terms of data collection, how would you avoid getting these types of observations?



---

# Sample variance and standard deviation
  
  * The mean and variance are the most popular
  
  * Nice mathematical properties
  
The sample variance, $s^2$:

$$s^2 = \frac{1}{n-1} \sum_{i=1}^n (x_i - \bar x)^2$$

--

The formula can be rewritten as

$$s^2= \frac{1}{(n-1)} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2 \right\}$$


---

# Sample variance and standard deviation
  

  * However $\sum x^2$ becomes very large as $n$ increases.
  
    - Loss of precision
    
  * In practice we would use
$$s^2= \sum_{i=1}^n \frac{x_i^2}{n-1}  - \left(\frac{n}{n-1}\right) \bar x^2$$


---

# Example

For our toy data set, $\{0,3,2,0\}$, we have

$$\sum_{i=1}^4 x_i^2 = 0^2 + 3^2 + 2^2 + 0^2 = 13$$

--

So
$$s^2 = \frac{1}{n-1} \left\{ \left(\sum_{i=1}^n x_i^2\right) - n \bar x^2 \right\} = \frac{1}{3} \left( 13 - 4 \times 1.25^2\right) = 2.25$$

--

 * The sample standard deviation, $s$, is the square root of the sample variance
  - $s = \sqrt{2.25} = 1.5$

  * Standard deviation is in the same units as the original data

---

# Rule of thumb

Approximately 95% of data is contained in the interval

$$\bar x \pm 2 \times s$$

> Tomorrow we'll see where this rule comes from

---

# Example: Beauty data set


```{r}
sd(beauty$evaluation)
var(beauty$evaluation)
```

---

# Quartiles and the interquartile range
  
  * $Q_3$: Upper quartile = $3 (n+1)/4^{\text{th}}$ smallest observation

  * $Q_1$: Lower quartile = $(n+1)/4^{th}$ smallest observation

$$IQR = Q_3 - Q_1$$

---

# R example

The quartiles are special cases of percentiles: 

 * The minimum is the $0^{th}$ percentile
 * $Q_1$ is the $25^{th}$ percentile
 * The median, $Q_2$ is the $50^{th}$ percentile
 * $Q_3$ is the $75^{th}$ percentile
 * The maximum is the $100^{th}$ percentile

```{r}
quantile(cupid$age)
```

---

# R example

```{r}
## The middle 90% 
quantile(cupid$age, probs = c(0.025, 0.975))
```

--

```{r}
# Rule of thumb
mean(cupid$age) - 2 * sd(cupid$age)
mean(cupid$age) + 2 * sd(cupid$age)
```


---

# Exercise

For the beauty data set calculate the `sd()`, `range()` and `quantiles()` for

  * `tenured`
  
  * `age`
  
  * `beauty`


---
`r add_border(inverse=TRUE, background_image = "assets/white_logo.png")`
---
`r add_class(inverse = TRUE, center = TRUE, middle = TRUE)`

## Streaming data is data that is generated continuously by multiple sources. 

---
`r add_border(inverse=FALSE)`
---

# The mean and variance

 * Suppose we have observed $k-1$ values and our current estimate of the mean is $m_{k-1}$
 
  - If it helps, just make $k$ a number, e.g. $k = 1000$

--

  * A new observation, $x_k$ arrives. How would we update our estimate? 
--

  * An obvious method is
$$m_k = \frac{1}{k} ((k-1) \times m_{k - 1} + x_k)$$

--

 * However this method isn't _numerically stable_ 

    * Basically $(k-1) \times m_{k - 1}$ gets large and we'll lose precision. 

---

# Streaming data: mean & variance

Instead we should use

$$m_k = m_{k-1} + \frac{x_k - m_{k-1}}{k}$$

We have a similar algorithm for the [variance](https://en.wikipedia.org/wiki/Algorithms_for_calculating_variance)

---

# Example: OKCupid

The OKCupid data set contains almost 60,000 individuals. 

```{r, echo=FALSE}
local(source("code/f2_average_age.R"))
```

---

# The median and quantiles

  * Keeping a [running median](https://stackoverflow.com/q/10657503/203420) is a non-trivial task
  
  * We need to maintain a sorted data structure containing the data
  
  * The key issues are storage cost of the data and retrieval time 

---

# Relevant R functions 

```{r echo=FALSE, results='asis'}
library(tibble)
r_functions = tribble(
  ~Command, ~Comment, ~Example, 
"mean", "Calculates the mean of a vector", "mean(x)",
"sd", "Calculates the standard deviation of a vector", "sd(x)",
"var", "Calculates the variance of a vector", "var(x)",
"quantile", "The vector quartiles" , "quantile(x)",
"range", "Calculates the vector range", "range(x)"
)  

knitr::kable(
  r_functions, longtable = TRUE, booktabs = TRUE,
  caption = 'Summary of R commands in this chapter.', format = "html"
)
```

    
